---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-",
  warning = FALSE, 
  message = FALSE,
  fig.align="center"
)
devtools::load_all()
```

# ebirdst: Tools to Load, Map, Plot, and Analyze eBird Status and Trends Data Products

<!-- [![License: GPL v3](https://img.shields.io/badge/License-GPL%20v3-blue.svg)](http://www.gnu.org/licenses/gpl-3.0) -->

**THIS PACKAGE IS UNDER ACTIVE DEVELOPMENT. FUNCTIONALITY MAY CHANGE AT ANY TIME.**

The goal of this package is to providet tools for loading, mapping, plotting, and analyzing eBird Status and Trends data products. If you're looking for the previous version (known as stemhelper), which is no longer being maintained, please access the stemhelper_archive branch.

## Installation

You can install ebirdst from GitHub with:

```{r gh-installation, eval = FALSE}
# install the development version from GitHub
# install.packages("devtools")
devtools::install_github("CornellLabofOrnithology/ebirdst")
```

## Vignette

TODO: update website, then updated this section.

For a full introduction and advanced usage, please see the package [website](https://cornelllabofornithology.github.io/ebirdst). An [introductory vignette](https://cornelllabofornithology.github.io/ebirdst/articles/stem-intro-mapping.html) is available, detailing the structure of the results and how to begin loading and mapping the results. Further, an [advanced vignette](https://cornelllabofornithology.github.io/ebirdst/articles/stem-pipd.html) details how to access additional information from the model results about predictor importance and directionality, as well as predictive performance metrics.

## Quick Start

This quick start guide will show you how to plot abundance values similar to how they are plotted for the [eBird Status and Trends Abundance animations](https://ebird.org/science/status-and-trends/woothr/abundance-map-weekly). First, you need to access and download the data.

TODO: Add details on how to acquire results.

**IMPORTANT. AFTER DOWNLOADING THE RESULTS AND UNZIPPING THEM, DO NOT CHANGE THE FILE STRUCTURE.** All functionality in this package relies on the structure inherent in the delivered results. Changing the folder and file structure will cause errors with this package. See the SETUP PATHS part of the example code below.

```{r quick_start}
library(viridis)
library(raster)
library(rnaturalearth)

# SETUP PATHS
# Once you have downloaded and unzipped the results, place the resulting folder,
# with the unique species ID name (e.g., woothr-ERD2016-PROD-20170505-3f880822)
# wherever you want to keep it. Then provide the root_path variable as that
# location (without the unique species ID name) and provide the unique species 
# ID name as the species variable below. Use the paste() function to combine

# This example downloads example data directly
#species <- "amerob-ERD2016-EXAMPLE_DATA-20171122-5a9e702e"
#species_url <- paste("http://ebirddata.ornith.cornell.edu/downloads/hidden/", 
#                     species, ".zip", sep = "")

#temp <- tempfile()
#temp_dir <- tempdir()
#download.file(species_url, temp)
#unzip(temp, exdir = temp_dir)
#unlink(temp)

#sp_path <- paste(temp_dir, "/", species, sep = "")

# Temp version before I put these on the web
# Loading data
root_path <- "~/Box Sync/Projects/2015_stem_hwf/documentation/data-raw/"
species <- "woothr-ERD2016-SP_TEST-20180724-7ff34421"
sp_path <- paste(root_path, species, sep = "")

# load Trimmed Mean Abundances and label
abunds <- raster::stack(paste0(sp_path, "/results/tifs/", species,
                               "_hr_2016_abundance_umean.tif"))
abunds <- label_raster_stack(abunds)

# to subset the data and add map context, get reference data from the rnaturalearth package
wh <- ne_countries(continent = c("North America"))
wh_states <- ne_states(iso_a2 = unique(wh@data$iso_a2))

# region for subsetting
wh_sub <- wh_states[wh_states$name %in% c("Minnesota", "Wisconsin", "Illinois", 
                                          "Iowa", "Indiana", "Ohio", "Michigan"), ]

# subset to a portion of the range
ne_extent <- list(type = "polygon",
                  polygon = wh_sub,
                  t.min = 0.48,
                  t.max = 0.50)
abund <- raster_st_subset(abunds, ne_extent)
rm(abunds)

# project to Mollweide for mapping and extent calc
mollweide <- CRS("+proj=moll +lon_0=-90 +x_0=0 +y_0=0 +ellps=WGS84")

# the nearest neighbor method preserves cell values across projections
abund_moll <- projectRaster(abund, crs = mollweide, method = 'ngb')

# project the reference data as well
wh_moll <- spTransform(wh, mollweide)
wh_states_moll <- spTransform(wh_states, mollweide)

# calculate ideal color bins for abundance values for this week
week_bins <- calc_bins(abund_moll)

# start plotting
par(mfrow = c(1, 1), mar = c(0, 0, 0, 6))

# use the extent object to set the spatial extent for the plot
plot(extent(trim(abund_moll, values = NA)), col = 'white')

# add background reference data
plot(wh_states_moll, col = "#eeeeee", border = NA, add = TRUE)

# plot zeroes as gray
plot(abund_moll == 0, maxpixels = ncell(abund_moll),
     ext = extent(trim(abund_moll, values = NA)), col = '#dddddd', 
     xaxt = 'n', yaxt = 'n', legend = FALSE, add = TRUE)

# define color bins
these_cols <- rev(viridis::plasma(length(week_bins$bins) - 2, end = 0.9))
grayInt <- colorRampPalette(c("#dddddd", these_cols[1]))
qcol <- c(grayInt(4)[2], these_cols)

# plot abundances
plot(abund_moll, maxpixels = ncell(abund_moll),
     ext = extent(trim(abund_moll, values = NA)), breaks = week_bins$bins,
     col = qcol, xaxt = 'n', yaxt = 'n', legend = FALSE, add = TRUE)

# for legend, create a smaller set of bin labels
bin_labels <- format(round(week_bins$bins, 2), nsmall = 2)
bin_labels[!(bin_labels %in% c(bin_labels[1],
                               bin_labels[round((length(bin_labels) / 2)) + 1],
                               bin_labels[length(bin_labels)]))] <- ""
bin_labels <- c("0", bin_labels)

# create colors that include gray for 0
lcol <- c('#dddddd', qcol)

# set legend such that color ramp appears linearly
ltq <- seq(from = week_bins$bins[1], to = week_bins$bins[length(week_bins$bins)],
           length.out = length(week_bins$bins))
ltq <- c(0, ltq)

# plot legend
plot(abund_moll ^ week_bins$power, col = lcol, legend.only = TRUE,
     breaks = ltq, lab.breaks = bin_labels, legend.shrink = 0.97,
     legend.width = 2, axis.args = list(cex.axis = 0.9, lwd.ticks = 0))

# add state boundaries on top
plot(wh_states_moll, add = TRUE, border = 'white', lwd = 1.5)

# clean up
#unlink(temp_dir)
```
