---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-",
  warning = FALSE, 
  message = FALSE
)
devtools::load_all()
```

# stemhelper: Helper functions for STEM loading, mapping, plotting, and analysis

<!-- [![License: GPL v3](https://img.shields.io/badge/License-GPL%20v3-blue.svg)](http://www.gnu.org/licenses/gpl-3.0) -->

**THIS PACKAGE IS UNDER ACTIVE DEVELOPMENT. FUNCTIONALITY MAY CHANGE AT ANY TIME.**

The goal of stemhelper is to provide functionality for loading, mapping, plotting, and analyzing STEM results.

## Installation

You can install stemhelper from GitHub with:

```{r gh-installation, eval = FALSE}
# install the development version from GitHub
# install.packages("devtools")
devtools::install_github("CornellLabofOrnithology/stemhelper")
```

## Vignette

For a full introduction and advanced usage, please see the package [website](https://cornelllabofornithology.github.io/stemhelper). An [introductory vignette](https://cornelllabofornithology.github.io/stemhelper/articles/stem-intro-mapping.html) is available, detailing the structure of the results and how to begin loading and mapping the results. Further, an [advanced vignette](https://cornelllabofornithology.github.io/stemhelper/articles/stem-pipd.html) details how to access additional information from the model results about predictor importance and directionality, as well as predictive performance metrics.

## Quick Start

TODO: Add details on how to acquire results.

**IMPORTANT. AFTER DOWNLOADING THE RESULTS AND UNZIPPING THEM, DO NOT CHANGE THE FILE STRUCTURE.** All functionality in this package relies on the structure inherent in the delivered results. Changing the folder and file structure will cause errors with this package. See the SETUP PATHS part of the example code below.

One of the most common first tasks is to load a RasterStack of relative abundance estimates and make a map.

```{r quick_start, fig.align = "center"}
library(viridis)
library(raster)
library(rnaturalearth)

# TODO set this up to do a curl of the example results first
# TODO make sure that the resulting unzipped folder is only single nested on both
# linux and Windows (had problems with norcar)

# SETUP PATHS
# Once you have downloaded and unzipped the results, place the resulting folder,
# with the unique species ID name (e.g., woothr-ERD2016-PROD-20170505-3f880822)
# wherever you want to keep it. Then provide the root_path variable as that
# location (without the unique species ID name) and provide the unique species 
# ID name as the species variable below. Use the paste() functiont to combine
root_path <- "~/Box Sync/Projects/2015_stem_hwf/documentation/data-raw/"
species <- "woothr-ERD2016-PROD-20170505-3f880822"
sp_path <- paste(root_path, species, sep = "")

# load a stack of rasters with the helper function stack_stem()
abund <- stack_stem(sp_path, variable = "abundance_umean")

# subset to the 26th week of the year
abund26 <- abund[[26]]

# project to Mollweide for mapping and extent calc
mollweide <- CRS("+proj=moll +lon_0=-90 +x_0=0 +y_0=0 +ellps=WGS84")
abund_moll <- projectRaster(abund26, crs = mollweide)

# calculate the full annual extent
# and set dimensions to set plotting size
sp_ext <- calc_full_extent(abund_moll)

xlimits <- c(sp_ext[1], sp_ext[2])
ylimits <- c(sp_ext[3], sp_ext[4])

xrange <- xlimits[2] - xlimits[1]
yrange <- ylimits[2] - ylimits[1]

# these are RMarkdown specific here, but could be passed to png()
w_cm <- 8.5
h_cm <- w_cm*(yrange/xrange)
knitr::opts_chunk$set(fig.width = w_cm, fig.height = h_cm)

# calculate ideal color bins for abundance values across the full year
year_bins <- calc_bins(abund_moll)

# to add context, pull in some reference data from the rnaturalearth package
wh <- ne_countries(continent = c("North America", "South America"))
wh_states <- ne_states(iso_a2 = unique(wh@data$iso_a2))
wh_moll <- spTransform(wh, mollweide)
wh_states_moll <- spTransform(wh_states, mollweide)

# start plotting
par(mfrow = c(1, 1), mar = c(0, 0, 0, 6))

# use the extent object to set the spatial extent for the plot
plot(sp_ext, col = 'white')

# add background spatial context
plot(wh_states_moll, col = "grey70", add = TRUE)

# plot the raster
plot(abund_moll,
     maxpixels = ncell(abund_moll),
     ext = sp_ext,
     breaks = year_bins,
     col = viridis(length(year_bins-1)),
     xaxt = 'n',
     yaxt = 'n',
     add = TRUE,
     legend = TRUE,
     legend.shrink = 0.97,
     legend.width = 2,
     axis.args = list(at = year_bins,
                      labels = round(year_bins, 2),
                      cex.axis = 0.7))

# add state boundaries on top
plot(wh_states_moll, add = TRUE, border = 'black', lwd = 1)
```
