---
output: md_document
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(
  warning = FALSE, 
  message = FALSE,
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-",
  fig.align = "center"
)
```

# ebirdst: Tools to Load, Map, Plot, and Analyze eBird Status and Trends Data Products

[![License: GPL v3](https://img.shields.io/badge/License-GPL%20v3-blue.svg)](http://www.gnu.org/licenses/gpl-3.0)
[![AppVeyor Build status](https://ci.appveyor.com/api/projects/status/v7cyxwquwrxxa5l6/branch/master?svg=true)](https://ci.appveyor.com/project/mstrimas/ebirdst/branch/master)
[![Travis build status](https://travis-ci.org/CornellLabofOrnithology/ebirdst.svg?branch=master)](https://travis-ci.org/CornellLabofOrnithology/ebirdst)
[![Coverage status](https://codecov.io/gh/CornellLabofOrnithology/ebirdst/branch/master/graph/badge.svg)](https://codecov.io/github/CornellLabofOrnithology/ebirdst?branch=master)

## Installation

You can install `ebirdst` from GitHub with:

```{r gh-installation, eval = FALSE}
# install the development version from GitHub
# install.packages("remotes")
remotes::install_github("CornellLabofOrnithology/ebirdst")
```

## Vignettes

For a full introduction and advanced usage, please see the package [website](https://cornelllabofornithology.github.io/ebirdst). An [introductory vignette](https://cornelllabofornithology.github.io/ebirdst/articles/ebirdst-introduction.html) details the data access and structure of the results. An [intro mapping vignette](https://cornelllabofornithology.github.io/ebirdst/articles/ebirdst-intro-mapping.html) expands upon the quick start readme and shows the basic mapping moves. The [advanced mapping  vignette](https://cornelllabofornithology.github.io/ebirdst/articles/ebirdst-advanced-mapping.html) shows how to reproduce the seasonal maps and statistics on the [eBird Status and Trends website](https://ebird.org/science/status-and-trends). Finally, the [non-raster data vignette](https://cornelllabofornithology.github.io/ebirdst/articles/ebirdst-pipd.html) details how to access additional information from the model results about predictor importance and directionality, as well as predictive performance metrics. 

## Quick Start

This quick start guide shows how to download example data and plot abundance values similar to how they are plotted for the [eBird Status and Trends Abundance animations](https://ebird.org/science/status-and-trends/woothr/abundance-map-weekly). **Important note: after downloading the results, do not change the file structure.** All functionality in this package relies on the structure inherent in the delivered results. Changing the folder and file structure will cause errors with this package. If you use this package to analyze the results, you do not ever need to interact with the files directly, outside of R.

```{r quick_start}
library(ebirdst)
library(viridis)
library(raster)
library(sf)
library(rnaturalearth)

# download data
# download a simplified example dataset from aws s3
# example data are for yellow-bellied sapsucker in michigan
# fby default ile will be stored in a persistent data directory:
# rappdirs::user_data_dir("ebirdst"))
sp_path <- ebirdst_download(species = "example_data")

# load estimated relative abundance and label with dates
# this raster stack has 52 layers, one for each week of the year
abunds <- load_raster("abundance_umean", path = sp_path)

# use parse_raster_dates() to get actual date objects for each layer
date_vector <- parse_raster_dates(abunds)
print(date_vector)

# select a week in the summer
abund <- abunds[[26]]
rm(abunds)

# project to mollweide for mapping
# the nearest neighbor method preserves cell values across projections
mollweide <- "+proj=moll +lon_0=-90 +x_0=0 +y_0=0 +ellps=WGS84"
abund_moll <- projectRaster(abund, crs = mollweide, method = "ngb")

# get reference data from the rnaturalearth package
# the example data currently shows only the US state of Michigan
wh_states <- ne_states(country = c("United States of America", "Canada"),
                    returnclass = "sf") %>% 
  st_transform(crs = mollweide) %>% 
  st_geometry()
  
# calculate ideal color bins for abundance values for this week
week_bins <- calc_bins(abund_moll)

# start plotting
par(mfrow = c(1, 1), mar = c(0, 0, 0, 6))

# use raster bounding box to set the spatial extent for the plot
bb <- st_as_sfc(st_bbox(trim(abund_moll)))
plot(bb, col = "white", border = "white")

# add background reference data
plot(wh_states, col = "#eeeeee", border = NA, add = TRUE)

# plot zeroes as gray
plot(abund_moll == 0, col = "#dddddd", 
     maxpixels = ncell(abund_moll),
     axes = FALSE, legend = FALSE, add = TRUE)

# define color bins
qcol <- abundance_palette(length(week_bins$bins) - 1, "weekly")

# plot abundances
plot(abund_moll, col = qcol, breaks = week_bins$bins,
     maxpixels = ncell(abund_moll),
     axes = FALSE, legend = FALSE, add = TRUE)

# for legend, create a smaller set of bin labels
bin_labels <- format(round(week_bins$bins, 2), nsmall = 2)
bin_labels[!(bin_labels %in% c(bin_labels[1],
                               bin_labels[round((length(bin_labels) / 2)) + 1],
                               bin_labels[length(bin_labels)]))] <- ""
bin_labels <- c("0", bin_labels)

# create colors that include gray for 0
lcol <- c("#dddddd", qcol)

# set legend such that color ramp appears linearly
ltq <- seq(from = week_bins$bins[1], to = week_bins$bins[length(week_bins$bins)],
           length.out = length(week_bins$bins))
ltq <- c(0, ltq)

# plot legend
plot(abund_moll ^ week_bins$power, legend.only = TRUE,
     col = lcol,
     breaks = ltq ^ week_bins$power, 
     lab.breaks = bin_labels, legend.shrink = 0.97,
     legend.width = 2, axis.args = list(cex.axis = 0.9, lwd.ticks = 0))

# add state boundaries on top
plot(st_geometry(wh_states), add = TRUE, col = NA, border = "white", lwd = 1.5)
```
