---
title: "Using Predictor Importance and Partial Dependence Information"
author: "Tom Auer, Daniel Fink"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using Predictor Importance and Partial Dependence Information}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, out.width="\\textwidth", 
                      fig.height=4, fig.width=7, fig.align="center")
devtools::load_all()
```

# Outline

1. Introduction
2. Loading Centroids
3. Selecting Region and Season
4. Plotting Centroids and Extent of Analysis
5. Plot Predictor Importance
6. Plot Partial Dependencies
7. Cake Plots

# 1. Introduction

To complete...

## Data Structure

Data for stixel centroids are stored in three files, found under \\results\\abund_preds\\unpeeled\_folds, as described below.

\\<species>
&nbsp;&nbsp;&nbsp;&nbsp;\\data
&nbsp;&nbsp;&nbsp;&nbsp;\\results
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;\\abund_preds
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;\\unpeeled_folds
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;\pi.txt
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;\pd.txt
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;\summary.txt

The stemhelper package provides functions for accessing these, such that you should never have to handle them manually, granted that the original file structure of the results is maintained.

# 2. Loading Centroids

```{r load_pipd}
# TODO set this up to do a curl of the example results first

# Note, if you currently access results you can use this path construction
# by replacing the root_path with where your copy lives
root_path <- "~/Box Sync/Projects/2015_stem_hwf/documentation/data-raw/"
species <- "woothr-ERD2016-PROD-20170505-3f880822"

# path to species bundle
sp_path <- paste(root_path, species, sep = "")

pis <- load_pis(sp_path)
pds <- load_pds(sp_path)
```

# 3. Selecting Region and Season

```{r map_centroids}
ne_extent <- list(type = "rectangle",
                  lat.min = 40,  
		              lat.max = 47,
		              lon.min = -80,
		              lon.max = -70,
		              t.min = 0.425,
	 	              t.max = 0.475)

map_centroids(pis = pis, pds = pds, st_extent = ne_extent)
```

# 4. Plotting Centroids and Extent of Analysis

```{r plot_extents}
par(mfrow = c(1,2), mar = c(0,0,0,6))
pi_extent <- calc_effective_extent(ne_extent, pis = pis)
pd_extent <- calc_effective_extent(ne_extent, pds = pds)
```

# 5. Predictive Performance Metrics

## Binary Metrics by Time

```{r binary_by_time}
plot_binary_by_time(sp_path, metric = "Kappa", n_time_periods = 52)
```

## All Metrics for Space-Time Extent

```{r all_ppms}
plot_all_ppms(sp_path, ne_extent)
```


# 6. Plot Predictor Importance

```{r plot_pis}
# with all classes
plot_pis(path = sp_path, pis, st_extent = ne_extent, by_cover_class = FALSE)

# aggregating fragstats for cover classes
plot_pis(path = sp_path, pis, st_extent = ne_extent, by_cover_class = TRUE)
```

# 7. Plot Partial Depedencies

```{r plot_pds}
mypds <- plot_pds(pd_name = "EFFORT_HRS", 
                  pds = pds, 
                  st_extent = ne_extent, 
                  pointwise_pi = TRUE, 
                  stixel_pds = TRUE)
```

# 8. Cake Plots

When PIs and PDs are combined with can get predictor importance with directionality. By week, these can be turned into a stacked area chart (we affectionately call cake plots) that show the relative predictor importance and directionality by week.

```{r cake_plots}
# this requires a fair amount of computation, about 5 minutes
cake_plot(path = sp_path,
          pis = pis,
          pds = pds,
          st_extent = ne_extent,
          by_cover_class = TRUE,
          pland_and_lpi_only = TRUE)
```
