---
title: "stemhelper: STEM Use and Helper Functions"
author: "Tom Auer & Daniel Fink"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE, out.width="\\textwidth", fig.height=4, 
                      fig.width=7, fig.align="center")
devtools::load_all()
```

# Outline

1. Background
2. Projections and Loading Rasters
3. Mapping Occurrence and Abundance
4. Extracting Trajectories with Uncertainty
5. Plotting Predictor Importance and Partial Dependencies
6. Conversion

# 1. Background

Many ecological and conservation problems require comprehensive, high-resolution information about the spatial and temporal distribution and abundance of species’ populations. This poses a challenge because observations of species are often sparsely and irregularly distributed, especially for widely distributed species. To address this challenge, species distribution models are used to estimate species’ distributions in unobserved locations and times. By associating local environmental characteristics with observed patterns of species occurrence and abundance statistical models provide a framework to harness available data and make estimates in unobserved areas. Additionally, these models can be used to control for known sources of biases that arise from the observation process.

The relative abundance estimates described here were based on bird observation data from [eBird](www.ebird.org), a citizen-science program administered by Cornell Lab of Ornithology, and local land cover descriptions derived from NASA MODIS and other remote sensing data. Individual species’ distributions and abundance were modeled using a spatio-temporal ensemble of boosted regression using the Spatio-Temporal Exploratory Model (STEM) framework (see Fink et al. 2010, 2014 and Johnston et al. 2015).

These models were then used to produce one daily abundance and occurrence estimate per week for all 52 weeks of the year. For each week, estimates were made across a regular grid of ~4 million locations that cover the terrestrial Western Hemisphere at a resolution of 2.8km x 2.8km. Variation in detectability associated with the search effort is controlled by standardizing the estimates to be the expected count of the species on a search conducted from 7-8AM while traveling 1 km on the given day at the given location by a typical eBirder.

Each relative abundance estimate was computed as an ensemble average of up to 100 randomized base models. Averaging across an ensemble of base model estimates, each trained on random subsample of the data, produces a smoothed bootstrap (Efron 2014) or a bagged estimator. For each abundance estimate we also computed upper and lower limits as the 10th and 90th percentiles across the ensemble. These limits are conservative estimates of the sampling variation of the smoothed bootstrap estimator, for the reasons discussed in Efron (2014), and because they are based on subsamples of the full data set.

## Data access

To access STEM results, begin by [creating an eBird account and signing in](https://secure.birds.cornell.edu/cassso/login). The visit the [Download Data](http://ebird.org/ebird/data/download) page. STEM data access is free; however, you will need to [request access](http://ebird.org/ebird/data/request) in order to obtain access to the STEM data. Filling out the access request form allows eBird to keep track of the number of people using the data and obtain information on the applications for which the data are used.

Once you have access to the data, proceed to the [download page](http://ebird.org/ebird/data/download/ebd). Here you will be able to download data for all species or for a list of 225 individual species. TBD...

## Data Structure and Types

Estimates from STEM are currently provided in the widely used GeoTiff raster format. These are easily opened with the raster package in R, as well as with a variety of GIS software tools. 

TODO: Revisit this section
The results are organized by species and each species has two groupings of results: "all\_layers" and "presentation". The "all\_layers" group contains the raw data layers about estimates and have not been thresholded. The "presentation" group contains data layers that have been thresholded. For each week, thresholded estimates have been restricted to locations that meet two criteria. First, each location was estimated to be occupied by the species, and, second, the location had sufficient eBird data to support non-zero abundance estimates for at least 50 of the 100 base models. 


The animated GIFs distributed with the data are generated using the "presentation" abundance estimates. It is highly suggested to use this GIF visualization as a means for sanity checking.

# 2. Projection and Loading Rasters

## Projection

The GeoTiffs use the same Sinusoidal projection that the MODIS data uses. As part of this package, we provide a template raster, that contains the spatial extent, resolution, and land water definition (as a 0 value) for our modeled region. To access, call the `template_raster` object after loading the package.

```{r show_template_raster, warning=FALSE, message=FALSE, cache=TRUE}
template_raster
```

## Loading Rasters

Once you have accessed and downloaded a set of STEM results, the first task is to load them into R in a way that they can be used easily. To save space on disk, each individual raster (e.g., abundance mean for Wood Thrush for January 4) has had NA values trimmed before saving (using the raster package trim() function). While this saves space on disk, it means that the rasters are not immediately comparable and trying to use the raster `stack()` function will fail. 

```{r load_single_raster, cache=TRUE}
# To load a single raster file and extend it to the full analysis extent
# Note that the template_raster object is included with this package
tif_path <- "~/Box Sync/Projects/2015_stem_hwf/documentation/data-raw/woothr/tifs/presentation/abundance_umean"

# TODO set this up to do a curl of the example results first
# change to a different variable
r <- raster::extend(raster::raster(paste(tif_path, "/", 
                                         list.files(tif_path)[1], 
                                         sep = "")),
                    template_raster)
r
```

To assist with the loading of rasters, we have provided a helper function, `stack_stem()`, which uses the `template_raster` object to extend each input raster in a directory to the full extent of analysis and then returns a `RasterStack` object.

```{r stack_abundance, cache=TRUE}
# TODO set this up to do a curl of the example results first

# load a stack of rasters with the helper function stack_stem()
abund_stack <- stack_stem(tif_path)
abund_stack
```

In most use cases, this function will be used to stack the 52 weeks of a single 
variable (e.g., abundance mean) for a single species. However, if the STEM .tif
rasters were re-arranged, the function could be used to, for example, stack 
single weeks for a single variable across multiple species.

# 3. Mapping Occurrence and Abundance

One of the most common activities with STEM results is to make plots of occurrence and abundance. It is fast and easy to make plots immediately, but to get ideal plotting behavior takes a little bit of preparation. This sections describes some helper functions intended to assist with mapping.

## Projections and Calculating Full-annual Spatial Extent

The sinusoidal projection that NASA provides MODIS data in, while functional because of its equal-area property, is not good for mapping because of the significant distortion of land masses across the globe. There are few global equal-area projections, including, but not limited to, Lambert Azimuthal Equal Area, Goode Homolosine, Mollweide, and Equal Area Cylindrical. Of these, we have selected Mollweide, used in the remainder of this section, for its pleasing conformal properties across the globe, ease of projection configuration within R and the proj4string construction, and the ease of setting central meridians that best display the Western Hemisphere. It is worth noting that while projecting to a more visually appealing projection is ideal for mapping, it is not necessary for quantitative analysis of STEM results, for which purposes the results can be left in the sinusoidal projection.

```{r project_stack, cache=TRUE}
library(viridis)

# define mollweide projection
mollweide <- sp::CRS("+proj=moll +lon_0=-90 +x_0=0 +y_0=0 +ellps=WGS84")

# project single layer from stack to mollweide
week1_moll <- raster::projectRaster(abund_stack[[1]], crs=mollweide)

# optionally, you can project an entire stack, but it takes much longer
#abund_stack_moll <- raster::projectRaster(abund_stack, crs=mollweide)

# plot single layer with full annual extent
par(mar=c(0,0,0,2))
raster::plot(week1_moll, 
             xaxt='n', 
             yaxt='n',
             bty='n',
             col=viridis(20),
             maxpixels=raster::ncell(week1_moll))
sp::plot(countries110_prj, add=TRUE)
```

Plotting the layer at the full spatial extent of the analysis area makes for a small map. The process of building the stack unifies the extent of all layers in the stack. However, this also means that plotting happens at the full spatial extent of the area of analysis, the Western Hemisphere, as seen above. If we calculate the full annual spatial extent for the species, project to Mollwide, and then map, we get an extent that works for the entire full-annual cycle of the species.

```{r plot_stack_extent, cache=TRUE}
# calculate the full annual extent on the abundance stack (in sinusoidal projection)
sp_ext <- calc_full_extent(abund_stack)

# project the full annual extent to Mollweide
extent_raster <- raster::raster(ext=sp_ext, 
                                crs=sp::proj4string(template_raster))
extent_raster[is.na(extent_raster)] <- 1
sp_ext_moll <- raster::extent(raster::projectRaster(extent_raster, 
                                                    crs=mollweide))

# plot the first week of the abundance stack in Mollweide
par(mar=c(0,0,0,2))
raster::plot(week1_moll, 
             xaxt='n', 
             yaxt='n', 
             bty='n',
             ext=sp_ext_moll, 
             col=viridis(20),
             maxpixels=raster::ncell(week1_moll))
```


## Mapping Occurrence

Most examples in this vignette focus primarily on abundance predictions, as they're the most novel and informative. However, we also provide the probability of occurrence predictions. These are much simpler to plot than abundance, as values range between 0 and 1 throughout the year.

```{r plot_occurrence, cache=TRUE}
# load a stack of occurrence
occ_path <- "~/Box Sync/Projects/2015_stem_hwf/documentation/data-raw/woothr/tifs/presentation/occurrence_umean"
occ_stack <- stack_stem(occ_path)

# create breaks every 0.05 from 0 to 1
occ_bins <- seq(0, 1, by = 0.05)
occ_week26 <- raster::projectRaster(occ_stack[[26]], crs=mollweide)

par(mar=c(0,0,0,2), cex=0.7)
raster::plot(occ_week26, 
             xaxt='n', 
             yaxt='n', 
             bty='n',
             ext=sp_ext_moll,
             breaks=occ_bins,
             col=viridis(length(occ_bins)-1),
             maxpixels=raster::ncell(occ_week26))

```


## Calculating Abundance Bins

Mapping species abundance across the full-annual cycle presents a challenge, in that patterns of concentration and dispersion in abundance change throughout the year, making it difficult to define color bins that suit all seasons and accurately reflect the detail of abundance predictions. To address this, we selected a method (described by Maciejewski et al. 2013) that log transforms the entire year of data, constructs bins with the log-transformed data using standard deviations, and then un-transforms the bins for plotting.

To compare, we first scale the colors linearly, based on the maximum from the entire year. 

```{r plot_linear, cache=TRUE}
year_max <- max(raster::maxValue(abund_stack), na.rm = TRUE)

week26_moll <- raster::projectRaster(abund_stack[[26]], crs = mollweide)

par(mfrow=c(1,2), mar=c(0,0,0,0))
raster::plot(week1_moll, 
             xaxt='n', 
             yaxt='n', 
             bty='n',
             ext=sp_ext_moll, 
             zlim=c(0, year_max),
             col=viridis(20),
             maxpixels=raster::ncell(week1_moll),
             legend=FALSE)
par(mar=c(0,0,0,4))
raster::plot(week26_moll, 
             xaxt='n', 
             yaxt='n', 
             bty='n',
             ext=sp_ext_moll, 
             zlim=c(0, year_max),
             col=viridis(20),
             maxpixels=raster::ncell(week26_moll))
```

We can compare this with plots made using the binning method provided here.

```{r plot_bins, cache=TRUE}
year_bins <- calc_bins(abund_stack)

# create a thinner set of labels
bin_labels <- format(round(year_bins, 2), nsmall = 2)
bin_labels[!(bin_labels %in% c(bin_labels[1], 
                    bin_labels[round((length(bin_labels)/4))+1], 
                    bin_labels[round((length(bin_labels)/2))+1],
                    bin_labels[length(bin_labels) - round((length(bin_labels)/4))+1],
                    bin_labels[length(bin_labels)]))] <- ""

par(mfrow=c(1,2), mar=c(0,0,0,0))
raster::plot(week1_moll, 
             xaxt='n', 
             yaxt='n', 
             bty='n',
             ext=sp_ext_moll, 
             breaks=year_bins,
             lab.breaks=bin_labels,
             col=viridis(length(year_bins)-1),
             maxpixels=raster::ncell(week1_moll),
             legend=FALSE)
par(mar=c(0,0,0,5))
raster::plot(week26_moll, 
             xaxt='n', 
             yaxt='n', 
             bty='n',
             ext=sp_ext_moll, 
             breaks=year_bins,
             lab.breaks=bin_labels,
             col=viridis(length(year_bins)-1),
             maxpixels=raster::ncell(week26_moll))
```

## Mapping Abundance

With a good full-annual extent and color bins, that accurately represent the data distribution, we can add in other pieces of data that represent predicted and assumed zeroes, which help display the extent of estimation.

```{r plot_es, cache=TRUE, warnings=FALSE}
# pull in species-specific ensemble support, represent predicted 0s where there is no abundance value
pos_es_path <- "~/Box Sync/Projects/2015_stem_hwf/documentation/data-raw/woothr/tifs/presentation/abundance_ensemble_support"
pos_es_stack <- stack_stem(pos_es_path)
week26_pes <- raster::projectRaster(pos_es_stack[[26]], crs=mollweide)

# pull in checklist-level ensemble support, represent assummed 0s
zero_es_path <- "~/Box Sync/Projects/2015_stem_hwf/documentation/data-raw/zero_es"
zero_es_stack <- stack_stem(zero_es_path)
week26_zes_vals <- raster::projectRaster(raster::mask(zero_es_stack[[26]], 
                                                 template_raster), 
                                    crs=mollweide)

# we'll combine the two types of zeroes with the positive prediction for plotting
week26_pes[week26_pes] <- 0
week26_zes <- week26_zes_vals >= 99
week26_zes[week26_zes == 0] <- NA
week26_zes[week26_zes == 1] <- 0

week26s <- raster::stack(week26_pes, week26_zes, week26_moll)
week26c <- raster::calc(week26s, max, na.rm=TRUE)

# to add context, let's pull in some reference data to add
wh <- rnaturalearth::ne_countries(continent = c("North America",
                                                "South America"))
wh_states <- rnaturalearth::ne_states(iso_a2 = unique(wh@data$iso_a2))
wh_moll <- sp::spTransform(wh, mollweide)
wh_states_moll <- sp::spTransform(wh_states, mollweide)

par(mar=c(0,0,0,2))
raster::plot(week26c, 
             xaxt='n', 
             yaxt='n',
             bty='n',
             ext=sp_ext_moll, 
             breaks=year_bins,
             lab.breaks=bin_labels,
             col=viridis(length(year_bins)-1),
             maxpixels=raster::ncell(week26c),
             legend=TRUE)
sp::plot(wh_moll, add=TRUE, border='gray')
sp::plot(wh_moll_states, add=TRUE, border='gray', lwd=0.5)
```

## Mapping Abundance Confidence Intervals

In addition to occurrence and abundance predictions, we also provide confidence intervals at an upper value of 90% and lower value of 10%. These can be used to calculate and map a confidence band width.

```{r plot_confidence, cache=TRUE}
# load lower and upper stacks
lower_path <- "~/Box Sync/Projects/2015_stem_hwf/documentation/data-raw/woothr/tifs/presentation/abundance_lower"
lower_stack <- stack_stem(lower_path)
upper_path <- "~/Box Sync/Projects/2015_stem_hwf/documentation/data-raw/woothr/tifs/presentation/abundance_upper"
upper_stack <- stack_stem(upper_path)

# calculate band width
conf_band <- upper_stack - lower_stack

conf_week26 <- raster::projectRaster(conf_band[[26]], crs=mollweide)

par(mar=c(0,0,0,2))
raster::plot(conf_week26, 
             xaxt='n', 
             yaxt='n',
             bty='n',
             ext=sp_ext_moll,
             col=magma(20),
             maxpixel=raster::ncell(conf_week26))

```

# 4. Extracting Trajectories with Uncertainty

With raster stacks for abundance, upper, and lower confidence intervals, we can extract an abundnace trajectory, with uncertainty intervals and plot them across the year for a single location.

```{r trajectories, cache=TRUE, warnings=FALSE}

# set a point
coords <- cbind(-76.5, 42.5)
pt <- sp::SpatialPoints(coords, proj4string = sp::CRS("+init=epsg:4326"))
pt_prj <- sp::spTransform(pt, sp::CRS(sp::proj4string(template_raster)))

# extract
abund_traj <- unlist(raster::extract(abund_stack, pt_prj, fun=mean, na.rm=TRUE))
upper_traj <- unlist(raster::extract(upper_stack, pt_prj, fun=mean, na.rm=TRUE))
lower_traj <- unlist(raster::extract(lower_stack, pt_prj, fun=mean, na.rm=TRUE))

# Plot trajectories
library(ggplot2)

plot_frame <- data.frame(x=1:length(abund_traj),
                         y=c(unname(abund_traj)),
                         upper=c(unname(upper_traj)),
                         lower=c(unname(lower_traj)))

g <- ggplot(plot_frame, aes(x, y)) +
     geom_line(data=plot_frame) +
     geom_ribbon(data=plot_frame, aes(ymin=lower, ymax=upper), alpha = 0.3) +
     ylab("Expected Count (count/(km hr))") +
     xlab("Week")
g

```

It is also possible to extract trajectories for regions, but it takes a little more data work.

```{r trajectories_region, cache=TRUE, warnings=FALSE}

# set a point
coords <- cbind(-76.5, 42.5)
pt <- sp::SpatialPoints(coords, proj4string = sp::CRS("+init=epsg:4326"))
pt_prj <- sp::spTransform(pt, sp::CRS(sp::proj4string(template_raster)))

# set an extent based on polygon
us <- rnaturalearth::ne_states(geounit = "United States of America")
us_prj <- sp::spTransform(us, sp::CRS(sp::proj4string(template_raster)))
wv <- us_prj[us_prj@data$name == "West Virginia",]

# extract
# because we're using regions, we get lots of values that we need to average together
abund_traj <- raster::extract(abund_stack, wv, fun=mean, na.rm=TRUE)
abund_traj <- apply(abund_traj, 2, mean, na.rm=TRUE)

upper_traj <- raster::extract(upper_stack, wv, fun=mean, na.rm=TRUE)
upper_traj <- apply(upper_traj, 2, mean, na.rm=TRUE)

lower_traj <- raster::extract(lower_stack, wv, fun=mean, na.rm=TRUE)
lower_traj <- apply(lower_traj, 2, mean, na.rm=TRUE)

# Plot trajectories
library(ggplot2)

plot_frame <- data.frame(x=1:length(abund_traj),
                         y=c(unname(abund_traj)),
                         upper=c(unname(upper_traj)),
                         lower=c(unname(lower_traj)))

g <- ggplot(plot_frame, aes(x, y)) +
     geom_line(data=plot_frame) +
     geom_ribbon(data=plot_frame, aes(ymin=lower, ymax=upper), alpha = 0.3) +
     ylab("Expected Count (count/(km hr))") +
     xlab("Week")
g

```

# 5. Plotting Predictor Importance and Partial Dependencies

Blah, blah, blah...

## Loading Data

```{r load_pipd, cache=TRUE}

# path to species bundle
sp_path <- "~/Box Sync/Projects/2015_stem_hwf/documentation/data-raw/woothr/"

pis <- load_pis(sp_path)
pds <- load_pds(sp_path)

```


## Select Region and Season

```{r plot_centroids, cache=TRUE, warning=FALSE, messages=FALSE}

ne_extent <- list()
ne_extent <- list(type = "rectangle",
                  y.min = 40,  
		              y.max = 47,
		              x.min = -80,
		              x.max = -70,
		              t.min = 0.425,
	 	              t.max = 0.475)

plot_centroids(pis=pis, pds=pds, st_extent=ne_extent)

```


## Plot available centroids and extent of analysis

```{r plot_extents, cache=TRUE}

pi_extent <- calc_effective_extent(ne_extent, pis = pis)

pd_extent <- calc_effective_extent(ne_extent, pds = pds)

```


## Plot Predictor Importance

## Plot Partial Depedencies


# 6. Conversion

The `raster` package has a lot of functionality and the RasterLayer format is useful for spatial analysis and mapping, but some users do not have GIS experience or want the data in a simpler format for their preferred method of analysis. There are multiple ways to get more basic representations of the data.

```{r conversion, warning=FALSE, cache=TRUE}
# To convert the data to a simpler geographic format and access tabularly	
# reproject into geographic (decimal degrees) 
abund_stack_ll <- raster::projectRaster(abund_stack[[26]], crs=sp::CRS("+init=epsg:4326"))

# Convert raster object into a matrix
p <- raster::rasterToPoints(abund_stack_ll)
head(p)
```

```{r conversion_write, warning=FALSE, eval=FALSE}
# Write a CSV file of the matrix
write.csv(p, 
          file = paste(raster_dir, "/woothr_week26.csv", sep=""),
          row.names = FALSE)
```

### References
